---
output: github_document
bibliography: inst/REFERENCES.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.height = 2.5, fig.width = 5, dpi = 150
)
```

# gasr

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/gasr)](https://CRAN.R-project.org/package=gasr)
<!-- badges: end -->

R functions for working with and simulating goal attainment scaling data.

## Installation

``` r
#install.packages("remotes")
remotes::install_github("taylordunn/gasr")
```

## Simulating data

The data simulation process is modularized as follows:

1. Simulate a number of subjects and assign to groups with `sim_subjects()`.
2. For each subject, simulate a number of goals with `sim_goals()`.
3. For each goal, simulate a treatment effect with `sim_treatment_effect()`.
4. For each goal, simulate a goal weight with `sim_goal_weights()`.
5. Get continuous latent goal scores by adding up the fixed and random effects.
6. Get discrete observed goal scores by applying thresholds with `discretize_from_thresholds()`.

For example, simulate 50 subjects, randomly allocated to "treatment" and "control" groups with equal probability:

```{r sim_subjects}
library(gasr)
library(tidyverse, quietly = TRUE)

gas_data <- sim_subjects(
  n_subjects = 50,
  group_allocation = list(control = 0.5, treatment = 0.5)
)
glimpse(gas_data)
```

Add between 3 and 6 goals to each subject:

```{r sim_goals}
gas_data <- gas_data %>%
  sim_goals(n_goals_range = c(3, 6))
glimpse(gas_data)
```

Note that `goals` is returned as a list column, so that the data frame keeps the same number of rows (one row per subject).
The goals can be unnested with `tidyr` to return a data frame with one row per goal:

```{r}
gas_data %>%
  unnest(goals) %>%
  glimpse()
```

A randomly sampled treatment effect can be added to each goal.
This function can be applied two ways.
Method A keeps the data nested, and applies the function to the `goals` list column using `purrr`:

```{r sim_treatment_effect_a}
set.seed(2)
gas_data_nested <- gas_data %>%
  mutate(
    goals = map(
      goals,
      ~sim_treatment_effect(., delta = 0.2) 
    )
  )
glimpse(gas_data_nested)
```

Method B is to `unnest`, `group_by` subject, and apply:

```{r sim_treatment_effect_b}
set.seed(2)
gas_data <- gas_data %>%
  unnest(goals) %>%
  group_by(subject_id) %>%
  sim_treatment_effect(delta = 0.2) %>%
  ungroup()
glimpse(gas_data)
```

Note that these two methods produce the same results because we gave them the same random seed:

```{r sim_treatment_effect}
bind_rows(
  A = gas_data_nested %>% unnest(goals),
  B = gas_data,
  .id = "method"
) %>%
  ggplot(aes(x = treatment_fe, y = method)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75))
```

Goal weights are applied the same way as treatment effects:

```{r sim_goal_weights1}
gas_data <- gas_data %>%
  group_by(subject_id) %>%
  sim_goal_weights(weight_type = "preference" ) %>%
  ungroup()
glimpse(gas_data)
```

There are a few different options of goal weights available, which are explained in the function documentation (`?sim_goal_weights`).
The "preference" weights here randomly apply integer weights from 1 to the number of goals per subject.
For instance, this subject with `r sum(gas_data$subject_id == "S01")` goals:

```{r sim_goal_weights2}
gas_data %>%
  filter(subject_id == "S01") %>%
  select(n_goals, goal_num, goal_weight)
```

The next step is to apply all of the fixed and random effects to get a total latent score for each goal:

```{r score_continuous, fig.height=2.5, fig.width=5}
gas_data <- gas_data %>%
  mutate(
    score_continuous = ifelse(group == "treatment", treatment_fe, 0) +
      subject_re + goal_re
  )
glimpse(gas_data)

gas_data %>%
  ggplot(aes(y = group, x = score_continuous, color = group)) +
  geom_jitter(width = 0, height = 0.2, alpha = 0.2) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.2, size = 1) +
  theme(legend.position = "none")
```

As a latent variable, this continuous score is not actually measured.
Rather a discretized goal score (usually a 5-point scale) is the observed variable.
Use the `create_thresholds()` function to create equally spaced thresholds on the latent variable score:

```{r create_thresholds}
thresh <- create_thresholds(n_levels = 5)
thresh
```

Then apply those weights manually with `cut()`, or with the convenience function `discretize_from_thresholds()`:

```{r score_discrete, fig.height=2.5, fig.width=5}
gas_data <- gas_data %>%
  mutate(
    score_discrete = discretize_from_thresholds(score_continuous, thresh)
  )
gas_data %>%
  ggplot(aes(x = score_discrete, fill = group)) +
  geom_bar(position = "dodge")
```

There is then a `calc_tscore()` function to compute an aggregate goal score based on the formula proposed by [@Kiresuk1968]:

```{r tscore, fig.height=2.5, fig.width=5}
gas_data <- gas_data %>%
  group_by(subject_id) %>%
  mutate(tscore = calc_tscore(score_discrete, goal_weight)) %>%
  ungroup()
gas_data %>%
  ggplot(aes(y = group, x = tscore, color = group)) +
  geom_jitter(width = 0, height = 0.2, alpha = 0.2) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.2, size = 1) +
  theme(legend.position = "none") +
  geom_vline(xintercept = 50)
```

All of these steps are packaged into a function with sensible defaults which simulates a single trial:

```{r sim_trial1}
d <- sim_trial()
glimpse(d)
```

Those defaults are listed below:

```{r sim_trial2}
formals(sim_trial)
```

## Simulation workflow

The recommended workflow for running many simulations of GAS trials is as follows:

1. Choose your trial parameters
    * Number of subjects `n_subjects`
    * Group allocation `group_allocation`
    * Treatment effect `delta`
2. Choose GAS parameters
    * Number of goals per subject: `n_goals`, `n_goals_range`, or `n_goals_prob`
    * Goal weights `weight_type`
3. Choose simulation parameters
    * Random seed via `set.seed()`
    * Number of trials to simulate `n_sim`

For the first two steps, the `sim_trial()` function can be used as a shortcut.
For instance, if we want to simulate 20 subjects with 3 goals and a treatment effect of 0.5:

```{r sim_trial3}
sim_trial(n_subjects = 20, n_goals = 3, delta = 0.5) %>%
  glimpse()
```

To incorporate the third step, it is recommended to write a custom function.
For example, if we just want to vary the number of subjects and the treatment effect (leaving all other options the same/default):

```{r run_simulations1}
run_simulations <- function(seed, n_sim, n_subjects, delta) {
  set.seed(seed)
  
  tibble(
    sim = 1:n_sim,
  ) %>%
    mutate(
      data = map(
        sim,
        ~sim_trial(n_subjects = n_subjects, delta = delta)
      ),
      seed = seed,
      n_subjects = n_subjects, delta = delta
    )
}
```

Then we can run 100 simulation with `n_subjects` = 20 and `delta` = 0.3 (and time it with the `tictoc` package):

```{r run_simulation2}
library(tictoc)
tic()
sims1 <- run_simulations(seed = 52, n_sim = 100,
                         n_subjects = 20, delta = 0.3)
toc()
```
The custom function returns a row per simulation, with the data nested in the `data` column.
Here is an example of unnesting and plotting the T-scores of the first 10 simulations:

```{r run_simulations3}
sims1 %>%
  unnest(data) %>%
  filter(sim <= 10) %>%
  distinct(sim, n_subjects, delta,
           subject_id, group, tscore) %>%
  ggplot(aes(x = tscore)) +
  geom_density(aes(fill = group), alpha = 0.5) +
  geom_vline(xintercept = 50, color = "white") +
  facet_wrap(~sim, nrow = 2) +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank())
```

## Todo

* Update documentation for `sim_trial`
* Continue `data-generating-model` vignette
* Write unit tests
* Write `power-calculation` vignette

## References
