---
title: "The GAS data generating model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The GAS data generating model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 3, fig.width = 5
)
```

```{r setup}
library(gasr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
```

With this package, we employ the latent variable approach to generating goal attainment scaling data that was introduced by [@Urach2019].
Here, we walk through some of the theory behind this model and show how it translates to the various `gasr` functions, but highly recommend any users to read the paper themselves to better understand the model.

First, some notation (which we alter slightly from that in @Urach2019).
Let $n_i \sim D$, where $D$ denotes the distribution of the number of goals chosen by subject $i$.
The observed goal score for subject $i$'s goal $j$ is denoted $x_{ij}$, and comes from the discretization of their continuous *latent* unobserved goal scores $y_{ij}$.
For goals $j = 1, \dots n_i$:

$$
\tag{1}
y_{ij} = b_0 + u_i + g_i b_{ij} + \epsilon_{ij}
$$

where

* $b_0$ is the mean response in the control group.
* $u_i \sim N(0, \sigma_u^2)$ is a patient-specific random effect.
    * Assume independent between subjects.
    * This introduces correlation between goals within a subject.
* $g_i$ is subject $i$'s treatment group.
    * $g_i$ = 0 for control and $g_i$ = 1 for experimental treatment.
* $b_{ij} \sim B$ is the effect of the experimental treatment on the attainment of goal $j$ for subject $i$.
    * Assume non-negative random variables with variance $\sigma_B^2$.
    * Assume independent between subjects.
    * Assume independent between goals of the same subjects.
    * This reflects the assumption that each subject chooses distinct goals that respond uniquely to the treatment.
* $\epsilon_{ij} \sim N(0, \sigma_{\epsilon}^2)$ is a random error term.

The within-subject correlation of two goal scores $y_{ij}$ and $y_{ij'}$ ($j \neq j'$) is given by:


$$
\tag{2}
\rho_g = \frac{\sigma_u^2}{\sigma_u^2 + g \sigma_B^2 + \sigma_{\epsilon}^2}
$$

which implies that, if $\sigma_B > 0$, the within-subject correlation in the treatment group is smaller than that in the control group.

<details><summary>Proof</summary>

The correlation between two random variables $X$ and $Y$ is:

$$
\rho = \frac{\text{Cov}(X, Y)}{\sqrt{\text{Var}(X) \text{Var}(Y)}}
$$

So if we consider the correlation between subject $i$'s two goals $j$ and $j'$:

$$
\tag{3}
\rho_g = \frac{\text{Cov}(y_{ij}, y_{ij'})}{\sqrt{\text{Var}(y_{ij}) \text{Var}(y_{ij'})}}
$$

Consider the variance of the latent goal score $\text{Var}(y_{ij})$.

$$
\text{Var}(y_{ij}) = \text{Var}(b_0 + u_i + g_i b_{ij} + \epsilon_{ij})
$$
To decompose the variance, we use the following theorem from @Casella2002:

>**Theorem 4.5.6** (p. 171) *If $X$ and $Y$ are any two random variables and $a$ and $b$ are any two constants, then*
$$
\text{Var}(aX + bY) = a^2 \text{Var} X + b^2 \text{Var} Y + 2ab \text{Cov}(X, Y)
$$

By construction, the four terms $b_0$, $u_i$, $b_{ij}$, and $\epsilon_{ij}$ are independent, and so $\text{Cov}(X, Y) = 0$ for each pair (**Thereom 4.5.5** (p. 171)):

$$
\begin{align}
\text{Var}(y_{ij}) &= \text{Var}(b_0) + \text{Var}(u_i) + g_i^2\text{Var}(b_{ij}) + \text{Var}(\epsilon_{ij}) \\
&= 0 + \sigma_u^2 + g_i^2 \sigma_B^2 + \sigma_\epsilon^2 \\
&= \sigma_u^2 + g \sigma_B^2 + \sigma_\epsilon^2 \\
\end{align}
$$

And the same is true for $\text{Var}(y_{ij'})$, of course.
So the denominator of equation (3) is:

$$
\tag{4}
\sqrt{\text{Var}(y_{ij}) \text{Var}(y_{ij'})} =
\sqrt{[\sigma_u^2 + g \sigma_B^2 + \sigma_{\epsilon}^2]^2} =
\sigma_u^2 + g \sigma_B^2 + \sigma_\epsilon^2 \\
$$

Now for the numerator of (3), the covariance between goals $j$ and $j'$.
From @Casella2002:

>**Definition 4.5.1** (p.169) *The covariance of $X$ and $Y$ is the number defined by*
$$
\text{Cov}(X, Y) = E((X - \mu_X)(Y - \mu_Y))
$$

Applying this to $y_{ij}$ and $y_{ij'}$:

$$
\tag{5}
\text{Cov}(y_{ij}, y_{ij'}) = E((y_{ij} - E(y_{ij}))(y_{ij'} - E(y_{ij'}))
$$

Consider just the first term of the product:

$$
\begin{align}
y_{ij} - E(y_{ij}) &= (b_0 + u_i + g_i b_{ij} + \epsilon_{ij}) - E(b_0 + u_i + g_i b_{ij} + \epsilon_{ij}) \\
&= (b_0 - b_0) + (u_i - E(u_i)) + g_i(b_{ij} - E(b_{ij})) + (\epsilon_{ij} - E(\epsilon_{ij})) \\
\end{align}
$$

And likewise for the term $y_{ij'} - E(y_{ij'})$.
Then the product in equation (5) is:

$$
\begin{align}
[y_{ij} - E(y_{ij})] [y_{ij'} - E(y_{ij'})] &=
[(u_i - E(u_i)) + g_i(b_{ij} - E(b_{ij})) + (\epsilon_{ij} - E(\epsilon_{ij}))]
[(u_i - E(u_i)) + g_i(b_{ij'} - E(b_{ij'})) + (\epsilon_{ij'} - E(\epsilon_{ij'}))] \\
&= (u_i - E(u_i))^2 + g_i^2(b_{ij} - E(b_{ij}))(b_{ij'} - E(b_{ij'})) + (\epsilon_{ij} - E(\epsilon_{ij}))(\epsilon_{ij'} - E(\epsilon_{ij'})) \\
&+ g_i(u_i - E(u_i))(b_{ij'} - E(b_{ij'})) + (u_i - E(u_i))(\epsilon_{ij'} - E(\epsilon_{ij'})) \\
&+ g_i(b_{ij} - E(b_{ij}))(u_i - E(u_i)) + g_i(b_{ij} - E(b_{ij}))(\epsilon_{ij'} - E(\epsilon_{ij'})) \\
&+ (\epsilon_{ij} - E(\epsilon_{ij}))(u_i - E(u_i)) + g_i (e_{ij} - E(\epsilon_{ij}))(b_{ij'} - E(b_{ij'}))
\end{align}
$$

Taking the expected values of the above gives us a sum of variance and covariance terms (by **Definition 4.5.1** above):

$$
\begin{align}
E[y_{ij} - E(y_{ij})] [y_{ij'} - E(y_{ij'})] &=
\text{Var}(u_i) + g_i^2 \text{Cov}(b_{ij}, b_{ij'}) + \text{Cov}(\epsilon_{ij}, \epsilon_{ij'}) \\
&+ g_i \text{Cov}(u_i, b_{ij'}) + \text{Cov}(u_i, \epsilon_{ij'}) \\
&+ g_i \text{Cov}(u_i, b_{ij}) + g_i\text{Cov}(b_{ij}, \epsilon_{ij'}) \\
&+ \text{Cov}(u_i, \epsilon_{ij}) + g_i \text{Cov}(b_{ij'}, \epsilon_{ij})
\end{align}
$$
As already established, we have independence between every pair combination of $b_0$, $u_i$, $b_{ij}$, $b_{ij'}$, $\epsilon_{ij}$, and $\epsilon_{ij'}$, and so all of the covariance terms are zero:

$$
\tag{6}
E[y_{ij} - E(y_{ij})] [y_{ij'} - E(y_{ij'})] = \text{Var}(u_i) = \sigma_u^2
$$

Combining the denominator from equation (4) and numerator from equation (6) into equation (3):

$$
\rho_g = \frac{\sigma_u^2}{\sigma_u^2 + g \sigma_B^2 + \sigma_{\epsilon}^2}
$$

gives us the expected expression of equation (2).

</details>

The discrete goal attainment levels $x_{ij}$ are defined by discretizing $y_{ij}$ via a set of thresholds:

$$
\tag{7}
-\infty = c_0 < c_1 \dots c_{2l + 1} = \infty
$$

such that $x_{ij} = t - l - 1$ if $c_{t-1} < y_{ij} \leq c_t$ for $t = 0, \dots, 2l+1$.
This means that $x_{ij}$ will range from $-l$ to $l$.

Traditionally, a 5-point attainment scale is employed in GAS.
This corresponds to $l = 2$, and $s = 1, \dots, 5$:

$$
\tag{8}
\begin{align}
t &= (0, \dots 2l + 1): c_{t-1} < y_{ij} \leq c_t \rightarrow x_{ij} = t - l - 1  \\
\\
t &= 0: \text{undefined } c_{-1} \\
t &= 1: c_0 < y_{ij} \leq c_1 \rightarrow x_{ij} = 1 - 2 - 1 = -2 \\
t &= 2: c_1 < y_{ij} \leq c_2 \rightarrow x_{ij} = 2 - 2 - 1 = -1 \\
t &= 3: c_2 < y_{ij} \leq c_3 \rightarrow x_{ij} = 3 - 2 - 1 = 0 \\
t &= 4: c_3 < y_{ij} \leq c_4 \rightarrow x_{ij} = 4 - 2 - 1 = +1 \\
t &= 5: c_4 < y_{ij} \leq c_5 \rightarrow x_{ij} = 5 - 2 - 1 = +2 \\
\end{align}
$$

(I'm not sure why $t = 0$ is considered, given that it is undefined).

## Simulation scenarios

In @Urach2019, the following simulation reference scenario is defined to compare hypothesis testing power:

* Parallel group design with equal per group sample size $m/2$.
* Overall sample size $m = 40$.
* Thresholds $c_t = \Phi^{-1} (0.2t)$, $t = 0, \dots, 5$.
    * $\Phi^{-1}$ denotes the inverse of the cumulative standard normal distribution function.

<details><summary>Thresholds</summary>

The cumulative standard normal probability distribution $\Phi$ can be computed with `stats::pnorm()`:

```{r fig.width=6}
d <- tibble(x = seq(-3, 3, 0.1)) %>%
  mutate(
    p = stats::pnorm(x, mean = 0, sd = 1),
  )

d %>%
  ggplot(aes(x, p)) +
  geom_line() +
  labs(title = "Cumulative standard normal distribution (mean 0, SD 1)")
```

The inverse cumulative probability function $\Phi^-1$ is also called the quantile function.
It returns the values of a random variable corresponds to percentiles:

```{r fig.width=6}
d <- d %>%
  mutate(
    q = stats::qnorm(p, mean = 0, sd = 1),
  )

d %>%
  ggplot(aes(p, q)) +
  geom_line() +
  scale_x_continuous(breaks = seq(0, 1.0, 0.2)) +
  geom_vline(xintercept = seq(0, 1, 0.2), lty = 2) +
  labs(title = "Inverse cumulative standard normal distribution (mean 0, SD 1)")
```

In the simulation scenario, six thresholds are defined:  $c_t = \Phi^{-1} (0.2t)$, $t = 0, \dots, 5$.
These are marked in the above plot with the dashed lines.
The values $c_t = \Phi^{-1}$ for those values of $t$:

```{r}
thresh <- tibble(
  t = 0:5,
  percentile = 0.2 * t,
  c_t = stats::qnorm(percentile, mean = 0, sd = 1) %>% round(2)
)
knitr::kable(thresh)
```

Six thresholds corresponds to five discretized attainment levels; $l = 2$ in equation (7).
To show how this works, consider these randomly sampled latent goal scores $y_{ij}$:

```{r fig.width=6}
y <- rnorm(1000, mean = 0, sd = 1)
ggplot(data = tibble(y), aes(y)) +
  geom_density() +
  geom_rug() +
  geom_vline(xintercept = thresh$c_t, lty = 2) +
  scale_x_continuous(limits = c(-5, 5))
```

Where the minimum and maximum thresholds at $\pm \infty$ were drawn at the edges of the plot, so they are visible.

We then bucket those scores according to to equation (8), for which `gasr` has a convenience function:

```{r}
x <- gasr::discretize_from_thresholds(y, thresholds = thresh$c_t)

ggplot(data = tibble(x), aes(x)) +
  geom_bar()
```

</details>

* The number of goals $n_i$ are uniformly distributed on $(1, \dots, n_{\text{max}}$) with $n_{\text{max}} = 5$.
* The treatment effects $b_{ij}$ are assumed to be uniformly distributed on $(0, 2\delta)$ for fixed constant $\delta$.
    * $B = U(0, 2\delta)$, $E(b_{ij}) = \delta$, $\text{Var}(b_{ij}) = \delta^{2/3}$.
* The mean response in the control group is set to zero.
    * $b_0 = 0$, $\mu_0 = 0$.
* Correlations fixed by setting the variance of the random effects:
    * $\sigma_u^2 = \rho_0$, and
    * $\sigma_{\epsilon}^2 = 1 - \rho_0$, therefore
    * equation (2) is $\rho_g = \rho_0 / (g\sigma_B^2 + 1)$.
        * $\rho_0$ in the control group, and
        * $\rho_1 = \rho_0 / (\sigma_B^2 + 1)$ in the treatment group.
        
<details><summary>Mean, variance, and covariance of the discretized goal scores</summary>

Figure 2(a) of @Urach2019 shows the relationship of of the expected value of the discretized goal attainment scores in the treatment group, $\mu_1 = E(x_{ij}(g = 1))$, and the standard deviation $\sigma$, versus treatment effect size $\delta$ for the reference scenario.
The expected value for the continuous scores is:

$$
\begin{align}
E(y_{ij}|g_i = 1) &= E(b_0) + E(\mu_i) + E(b_{ij}) + E(\epsilon_{ij}) \\
&= 0 + 0 + \delta + 0 \\
&= \delta
\end{align}
$$

and the variance of $y_{ij}$ can be decomposed into:

$$
\begin{align}
\text{Var}(y_{ij}) = \sigma^2 &= \text{Var}(b_0) +\text{Var}(u_i) + \text{Var}(b_{ij}) + \text{Var}(\epsilon_{ij}) \\
&= 0 + \sigma_u^2 + \delta^{2/3} + \sigma_{\epsilon}^2 \\
&= \rho_0 + \delta^{2/3} + (1 - \rho_0) \\
&= \delta^{2/3} + 1
\end{align}
$$
The expected value of the discrete scores are more complicated.
Derivations are provided by @Urach2019 in Appendix 1, which we will attempt to work through here.

Using the cumulative normal distribution $\Phi$ defining the thresholds, they show that the marginal probability distribution of $x_{ij}$ is given by:

$$
\tag{9}
P(x_{ij} = t | g_i = g, b_{ij} = b) = \Phi\left(\frac{c_t - gb - b_0}{\sigma_u^2 + \sigma_e^2}\right) - \Phi \left(\frac{c_{t-1} - gb + b_0}{\sigma_u^2 + \sigma_e^2} \right)
$$

The moments of a distribution are defined in @Casella2002 as:

>**Definition 2.3.1** (p. 59) *For each integer $n$, the $n$th moment of $X$, $\mu_n'$ is*
$$
\mu_n' = E(X^n).
$$
*The $n$th central moment of $X$, $\mu_n$ is*
$$
\mu_n = E(X - \mu)^n,
$$
*where $\mu = \mu_1' = EX$.*

For a random variable $g(X)$ with probability distribution function $f_X(x)$, the integral definition of the expected value is:

>**Definition 2.2.1** (p. 55) *The expected value or mean of a random variable $g(X)$ is*
$$
\begin{align}
E(X) &= \int_{-\infty}^{\infty} g(x) f_X(x) dx \ \ \ \ \ \text{if X is continuous} \\
&= \sum_{x \in \mathcal{X}} g(x) f_X(x) \ \ \ \ \ \ \ \ \ \ \ \text{if X is discrete} \\
\end{align}
$$

With $f_b$ representing the density of $b_{ij}$, the non-central moments of $x_{ij}$ for $a >0$ are:

$$
\tag{10}
\begin{align}
E(x_{ij}^a | g_i = g) &= \int_B E(x_{ij}|g_i = g, b_{ij} = b) f_b (b) db \\
&= \sum_{t \in \mathcal{T}} t^a \int_B P(x_{ij} = t | g_i = g, b_{ij} = b) f_b(b) db
\end{align}
$$

I don't quite follow this, but the central moments are then given by:

$$
\begin{align}
\mu_g &= E(x_{ij}|g_i = g) \\
\sigma^2 = \text{Var}(x_{ij}|g_i = g) &= E(x_{ij}^2|g_i = g) - E(x_{ij}|g_i = g)^2
\end{align}
$$
(unless I'm misunderstanding, $\mu_g$ is not a central moment, it is the first moment as per the @Casella2002 definition).

Plugging equation (9) into (10) for $a = 1$, and using $f_b = 1/2\delta$ for the uniform treatment effect distribution:

$$
\begin{align}
\tag{11}
E(x_{ij} | g_i = g)
&= \sum_{t \in \mathcal{T}} t \int_B P(x_{ij} = t | g_i = g, b_{ij} = b) f_b(b) db \\
&= \sum_{t \in \mathcal{T}} t \int_B \left[\Phi\left(\frac{c_t - gb - b_0}{\sigma_u^2 + \sigma_e^2}\right) - \Phi \left(\frac{c_{t-1} - gb + b_0}{\sigma_u^2 + \sigma_e^2} \right)  \right] \frac{1}{2\delta} db
\end{align}
$$

We can simplify the cumulative distribution functions using the other assumptions from the reference scenario:

$$
\begin{align}
\Phi \left(\frac{c_t - gb - b_0}{\sigma_u^2 + \sigma_{\epsilon}^2}\right)
= \Phi \left(\frac{c_t - gb - 0}{\rho_0 + 1 - \rho_0} \right) = \Phi (c_t - gb)
\end{align}
$$

Then applying the formula for the standard normal CDF from equation (3.3.13) on p. 102 of @Casella2002:

$$
\Phi(c_t - gb) = \frac{1}{2\pi} \int_{-\infty}^x e^{-(c_t - gb)^2/2} db
$$

Note that this integration cannot be performed directly (see p. 103 of @Casella2002 for details).
We can numerically compute these values, however.
Below shows $\Phi(c_t - b)$ for different values of $t$:

```{r fig.width=6}
c_t <- stats::qnorm(seq(0, 1, 0.2), mean = 0, sd = 1)
d <- tibble(
  t = 1:5,
  t_label = str_c("t = ", t),
  c_t = c_t[2:6]
) %>%
  crossing(b = seq(-3, 3, 0.01), g = c(0, 1)) %>%
  mutate(phi = pnorm(c_t - g * b, mean = 0, sd = 1))
d %>%
  ggplot(aes(x = b, y = phi)) +
  geom_line(aes(color = factor(g))) +
  facet_wrap(~t_label)
```

And then we take the difference in adjacent distributions, which are shown below for the treatment group ($g = 1$):

```{r fig.height=3, fig.width=6}
c_t <- stats::qnorm(seq(0, 1, 0.2), mean = 0, sd = 1)
calc_phi_diff <- function(b, t) {
  pnorm(c_t[t + 1] - b, mean = 0, sd = 1) -
    pnorm(c_t[t] - b, mean = 0, sd = 1)
}

tibble(t = 1:5) %>%
  crossing(b = seq(-3, 3, 0.01)) %>%
  mutate(
    phi_diff = calc_phi_diff(b, t),
    t_label = str_c("phi(t = ", t + 1, ") - phi(t = ", t, ")")
  ) %>%
  ggplot(aes(x = b, y = phi_diff)) +
  geom_line(aes(color = t_label)) +
  scale_color_viridis_d()
```


And then, we can `integrate()` equation (11) for different values of $t$ and $delta$ to get the expected values:

```{r fig.width=5}
integrate_phi <- function(t, delta) {
  stats::integrate(
    f = function(b) calc_phi_diff(b, t),
    lower = 0, upper = 2 * delta, subdivisions = 100
  )$value
}

phi_integral <- tibble(t = as.integer(1:5)) %>%
  crossing(delta = seq(0.01, 5, 0.01)) %>%
  rowwise() %>%
  mutate(
    integral = integrate_phi(t, delta)
  ) %>%
  ungroup()

# Compute moments of the discrete goal scales by summing over t
x_moments <- phi_integral %>%
  group_by(delta) %>%
  summarise(
    integral_sum1 = sum((t - 3) * integral) / (2 * delta),
    integral_sum2 = sum((t - 3)^2 * integral) / (2 * delta),
    .groups = "drop"
  ) %>%
  mutate(
    mu = integral_sum1,
    var = integral_sum2 - mu^2
  )
x_moments %>%
  transmute(delta, mu, sigma = sqrt(var)) %>%
  pivot_longer(c(mu, sigma), names_to = "moment", values_to = "value") %>%
  ggplot(aes(x = delta, y = value, color = moment)) +
  geom_line() +
  scale_y_continuous(limits = c(0, 2))
```

The shape of the curves looks close to Figure 2(a), but I haven't quite gotten the scale right.

</details>

## Simulation examples

`gasr` provides modularized functions for running simulations with the specified parameters in the reference scenario.
For instance, if we let $\rho_0 = 0$, simulate 40 subjects equally allocated to control and treatment groups:

```{r}
set.seed(77)

rho_0 <- 0
var_u <- rho_0
var_e <- 1 - rho_0

gas_data <- sim_subjects(
  n_subjects = 40, sigma_u = sqrt(var_u),
  group_allocation = list(control = 0.5, treatment = 0.5),
  random_allocation = FALSE
)
gas_data %>% glimpse()
```

Give each subject between 1 and 5 (=$n_{\text{max}}$) goals:

```{r}
gas_data <- gas_data %>%
  sim_goals(n_goals_range = c(1, 5), sigma_e = sqrt(var_e)) %>%
  unnest(goals)
glimpse(gas_data)
```

Assign a uniform distribution treatment effect with $\delta$ = 0.5.

```{r}
gas_data <- gas_data %>%
  sim_treatment_effect(delta = 0.5)
glimpse(gas_data)
```

The goal weights could be added here with `sim_goal_weights`, but we'll leave them unweighted for this example.

Calculate the latent goal scores by summing up the random effects (`subject_re` = $u_i$, `goal_re` = $\epsilon_{ij}$) and fixed effects (`treatment_fe` = $b_{ij}$, no control group response $b_0$):

```{r}
gas_data <- gas_data %>%
  mutate(
    score_continuous = subject_re + goal_re +
      ifelse(group == "treatment", treatment_fe, 0)
  )
```

The distribution of continuous scores by treatment group:

```{r }
gas_data %>%
  ggplot(aes(x = score_continuous, fill = group, color = group)) +
  geom_density(alpha = 0.5) +
  geom_rug()
```

Now to discretize the scores, we define the thresholds using the inverse cumulative normal distribution:

```{r}
thresh <- create_thresholds(n_levels = 5)
thresh
```

And apply to the continuous scores:

```{r}
gas_data <- gas_data %>%
  mutate(
    score_discrete = discretize_from_thresholds(score_continuous, thresh)
  )
gas_data %>%
  ggplot(aes(x = score_discrete, fill = group)) +
  geom_bar(position = position_dodge())
```

It is helpful to run these functions step-by-step to understand the simulations, but it is not very efficient.
The `sim_trial()` function does all of the steps in one function:

```{r}
set.seed(15)

gas_data <- sim_trial(
  n_subjects = 40, sigma_u = sqrt(var_u),
  group_allocation = list("control" = 0.5, "treatment" = 0.5),
  random_allocation = FALSE,
  n_goals_range = c(1, 5), sigma_e = sqrt(var_e),
  delta = 0.5,
  n_levels = 5
)
glimpse(gas_data)
```

(Note that this function also returns `goal_weight`s, all equal to 1, and subject-level summary `tscore`s).

```{r}
gas_data %>%
  ggplot(aes(x = score_discrete, fill = group)) +
  geom_bar(position = position_dodge())
```

A simple hypothesis test would be to compare the subject-level mean goal attainment $\bar{x}_i = \frac{1}{n_i} \sum x_{ij}$ in the two groups via a Welch's $t$-test (allowing for unequal variance):

```{r}
gas_data_mean <- gas_data %>%
  group_by(subject_id, group) %>%
  summarise(
    mean_score_discrete = mean(score_discrete),
    .groups = "drop"
  )
t.test(
  mean_score_discrete ~ group,
  data = gas_data_mean,
  alternative = "two.sided", var.equal = FALSE
)
```

In this case, the difference was not statistically significant at $\alpha = 0.05$.

We can repeat this, say, 10000 times, run $t$-tests, and get the $p$ values:

```{r eval=FALSE}
set.seed(21)

gas_sims <-
  tibble(
    sim = 1:10000,
  ) %>%
  mutate(
    data = map(
      sim,
      ~sim_trial(
        n_subjects = 40, sigma_u = sqrt(var_u),
        group_allocation = list("control" = 0.5, "treatment" = 0.5),
        random_allocation = FALSE,
        n_goals_range = c(1, 5), sigma_e = sqrt(var_e),
        delta = 0.5,
        n_levels = 5
      ) %>%
        group_by(subject_id, group) %>%
        summarise(mean_score_discrete = mean(score_discrete),
                  .groups = "drop")
    )
  )
# Run t-tests on every simulation
gas_sims_ttest <- gas_sims %>%
  transmute(
    sim,
    ttest = map(
      data,
      ~t.test(mean_score_discrete ~ group,
              data = .,
              alternative = "two.sided", var.equal = FALSE)
    )
  ) %>%
  mutate(ttest = map(ttest, broom::tidy)) %>%
  unnest(ttest)
```

(This took about 8.5 minutes to run, for reference).

```{r echo=FALSE}
# To save time while building the vignette, the results of the simulation
#  were saved separately
gas_sims_ttest <- readr::read_rds(
  system.file("extdata/gas-sims-ttest.rds", package = "gasr")
)
```

```{r}
glimpse(gas_sims_ttest)
```

A total of `r sum(gas_sims_ttest$p.value < 0.05)` of these simulations (`r scales::percent(mean(gas_sims_ttest$p.value < 0.05))`) returned a $p$ value <0.05.
And so the power to detect a significant treatment effect in this scenario is 55% when using a $t$-test on per-subject means (which matches the power reported by @Urach2019).

## References
