---
title: "Internal consistency of goal scores"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Internal consistency of goal scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 3, fig.width = 5
)
```

```{r setup}
library(gasr)
library(dplyr)
```

A key metric in the theory of GAS, and in psychometrics in general, is *internal consistency*.
This concept is discussed in Chapter 9 (p. 192) of @Kiresuk1994, where they cite the following definition:

>Internal consistency refers to an estimate of test reliability based upon the average intercorrelation of items within the test [@Nunnally1978].

>The average scale intercorrelation can be estimated most easily by the intraclass correlation obtained from a subjects-by-scales analysis of variance.

We've [shown elsewhere](data-generating-model.html) that the ICC of continuous goal scores $y_{ij}$ and $y_{ij'}$ is a function of treatment group:

$$
\rho_g = \frac{\sigma_u^2}{\sigma_u^2 + g \sigma_B^2 + \sigma_{\epsilon}^2}.
$$

To compute this empirically, @Kiresuk1994 give the following formula:

$$
\tag{9.7}
\bar{r}_{ij} = \frac{\text{MS}_B - \text{MS}_W}{\text{MS}_B + (k - 1) \text{MS}_W},
$$

where $\bar{r}_{ij}$ is the average scale intercorrelation, $\text{MS}_B$ is the between-subjects mean square, $\text{MS}_W$ is the within-subjects mean square, and $k$ is the number of scales (goals) per subject.

<details><summary>Proof</summary>

For $n$ subjects with $k$ goals each and no treatment effect, the between-subjects mean square (on the latent score $y_{ij}$) is

$$
\text{MS}_B = k\sigma_u^2 + \sigma_e^2
$$

with degrees of freedom $df = n-1$, where $\sigma_u^2$ is the subject-level random effect variance, and $\sigma_{\epsilon}^2$ is the goal-level random effect variance.

The within-subjects mean square is

$$
\text{MS}_W = \sigma_{\epsilon}^2
$$

with degrees of freedom $df = n(k-1)$.

Rearranging $\text{MS}_B$ in terms of the subject-level variance:

$$
\sigma_u^2 = \frac{\text{MS}_B - \sigma_{\epsilon}^2}{k} = \frac{\text{MS}_B - \text{MS}_W}{k}
$$

Then Pearson's correlation is calculated as:

$$
\begin{align}
\bar{r}_{ij} &= \frac{\sigma_u^2}{\sigma_u^2 + \sigma_{\epsilon}^2} \\
&= \frac{\frac{\text{MS}_B - \text{MS}_W}{k}}{\frac{\text{MS}_B - \text{MS}_W}{k} + \text{MS}_W} \\
&= \frac{\text{MS}_B - \text{MS}_W}{\text{MS}_B + (k - 1) \text{MS}_W}
\end{align}
$$

</details>

For instance, simulate 20 subjects in a control group with three goals each:

```{r}
gas_data <- sim_trial(n_subjects = 20, n_goals = 3,
                      group_allocation = list("control" = 1.0))
```

Manually compute the total between-subjects sum of squares (SS) and mean:

```{r}
gas_data_between <- gas_data %>%
  mutate(grand_mean = mean(score_continuous)) %>%
  group_by(grand_mean, subject_id) %>%
  summarise(subject_mean = mean(score_continuous),
            .groups = "drop") %>%
  summarise(
    ss_b = sum((subject_mean - grand_mean)^2),
    df = n() - 1,
    ms_b = ss_b / df
  )
gas_data_between
```

And within-subjects:

```{r}
gas_data_within <- gas_data %>%
  group_by(subject_id) %>%
  summarise(
    ss_subj = sum((score_continuous - mean(score_continuous))^2),
    .groups = "drop"
  ) %>%
  summarise(
    ss_w = sum(ss_subj),
    df = n() * (3 - 1), # n*(k-1)
    ms_w = ss_w / df
  )
gas_data_within
```

Then the inter-correlation is:

```{r}
(gas_data_between$ms_b - gas_data_within$ms_w) /
  (gas_data_between$ms_b + (3 - 1) * gas_data_within$ms_w)
```



$$
\bar{r}_{ij} = \frac{\text{MS}_B - \text{MS}_W}{\text{MS}_B + (k-1) \text{MS}_W}
$$

```{r}

```


```{r}
aov(score_continuous ~ subject_id, data = gas_data) %>%
  summary()
```


```{r}
gas_data
```



When the number of 

@Streiner2008

## References
